---
title: "PFlogPF normalization in Seurat with scclrR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{PFlogPF normalization in Seurat with scclrR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

`scclrR` is designed to feel like a normal Seurat preprocessing step. It reads
a feature-by-cell count matrix from a Seurat assay, computes PFlogPF in Rust, and
writes two pieces back to the object:

1. a sparse assay layer containing the shifted log values, and
2. a per-cell centering vector in object metadata.

The normalized value used for PCA is the implicit matrix
`sparse_layer - center_vector`, so the centered CLR matrix does not need to be
stored densely.

## A Seurat object

This vignette uses Seurat's bundled PBMC example when available. For benchmark
work, the same calls apply to the Ahlmann-Eltze-Huber count matrices after they
are read into a Seurat object with features in rows and cells in columns.

```{r setup, eval = FALSE}
library(scclrR)
library(Seurat)

data("pbmc_small")
pbmc <- pbmc_small
DefaultAssay(pbmc) <- "RNA"
```

For an external Matrix Market file, the typical Seurat setup is:

```{r external-example, eval = FALSE}
counts <- Matrix::readMM("counts.mtx.gz")
genes <- read.delim("features.tsv.gz", header = FALSE)
cells <- read.delim("barcodes.tsv.gz", header = FALSE)
rownames(counts) <- make.unique(genes[[1]])
colnames(counts) <- cells[[1]]

obj <- CreateSeuratObject(counts = counts, assay = "RNA")
```

## Normalize with PFlogPF

`target = "auto"` estimates the overdispersion `alpha` and sets
`K = 4 * alpha * mean_depth`. You can also pass a numeric `target` to use a
fixed `K`, or pass `alpha = ...` directly.

```{r normalize, eval = FALSE}
pbmc <- pflogpf(
  pbmc,
  assay = "RNA",
  layer = "counts",
  target = "auto",
  key.added = "pflogpf"
)

pbmc@misc$scclrR$pflogpf
head(pbmc[["pflogpf_center"]])
```

The sparse shifted-log values are stored in an assay layer named `"pflogpf"`.
The cell centering vector is stored in metadata as `"pflogpf_center"`.

## PCA without densifying

`run_pca()` reads both pieces and calls the Rust `rupca` backend on the implicit
shifted-CLR matrix.

```{r pca, eval = FALSE}
pbmc <- run_pca(
  pbmc,
  assay = "RNA",
  layer = "pflogpf",
  center.key = "pflogpf_center",
  reduction.name = "scclr_pca",
  n.components = 10
)

Embeddings(pbmc, "scclr_pca")[1:5, 1:5]
Loadings(pbmc, "scclr_pca")[1:5, 1:5]
```

The resulting reduction can be used wherever Seurat expects a dimensional
reduction:

```{r downstream, eval = FALSE}
pbmc <- FindNeighbors(pbmc, reduction = "scclr_pca", dims = 1:10)
pbmc <- FindClusters(pbmc)
pbmc <- RunUMAP(pbmc, reduction = "scclr_pca", dims = 1:10)
```

## Representation details

For a count matrix with genes in rows and cells in columns, `scclrR` passes the
`dgCMatrix` slots directly to Rust. The column-compressed representation of a
gene-by-cell matrix is exactly the row-compressed representation of a
cell-by-gene matrix, so the Rust code can operate in the same orientation used
by `scclr` for AnnData.

The dense value for cell `i`, gene `j` is

```text
log_values[i, j] - center[i]
```

where `log_values` is sparse and `center` is dense of length `ncol(counts)`.
This is why PFlogPF can remain sparse for storage while still using the exact
centered geometry for PCA.
