---
title: "PFlogPF PCA for Angelidis lung pseudobulk RNA-seq"
output:
  rmarkdown::html_vignette:
    fig_width: 7
    fig_height: 5
vignette: >
  %\VignetteIndexEntry{PFlogPF PCA for Angelidis lung pseudobulk RNA-seq}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
has_deps <- all(vapply(
  c("Seurat", "SeuratObject", "edgeR", "ggplot2", "ggrepel"),
  requireNamespace,
  quietly = TRUE,
  FUN.VALUE = logical(1)
))
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  eval = has_deps
)
```

This vignette reproduces the first idea in manuscript Figure 3 with the R API:
PFlogPF can be useful not only for individual-cell matrices, but also for
sample-level pseudobulk RNA-seq. The example uses mouse-level lung pseudobulk
counts from Angelidis et al. 2019, included in `inst/extdata`.

`scclrR` stores PFlogPF as two pieces:

- a sparse shifted-log matrix, stored as a Seurat assay layer when the installed
  Seurat version supports layers, and
- a per-sample centering vector, stored in Seurat metadata.

PCA is then run on the implicit centered matrix without densifying it.

```{r setup}
library(scclrR)
library(Matrix)
library(Seurat)
library(edgeR)
library(ggplot2)
library(ggrepel)
```

## Read the included pseudobulk data

The count matrix is genes by mouse samples, which is the orientation expected by
Seurat.

```{r read-data}
counts <- readMM(system.file(
  "extdata",
  "angelidis_lung_pseudobulk_counts.mtx.gz",
  package = "scclrR"
))
samples <- read.csv(system.file(
  "extdata",
  "angelidis_lung_pseudobulk_samples.csv",
  package = "scclrR"
))
genes <- read.csv(system.file(
  "extdata",
  "angelidis_lung_pseudobulk_genes.csv",
  package = "scclrR"
))

rownames(counts) <- make.unique(genes$gene)
colnames(counts) <- samples$sample

lung <- CreateSeuratObject(counts = counts, meta.data = samples)
DefaultAssay(lung) <- "RNA"
table(lung$age_label)
```

## PFlogPF normalization and PCA

`target = "auto"` estimates the overdispersion and sets the shifted-log scale
`K = 4 * alpha * mean_depth`. The returned object contains a sparse PFlogPF
layer and a metadata column named `pflogpf_center`.

```{r pflogpf}
lung <- pflogpf(
  lung,
  assay = "RNA",
  layer = "counts",
  target = "auto",
  key.added = "pflogpf"
)

lung@misc$scclrR$pflogpf[c("k", "alpha", "center_key")]
head(lung[["pflogpf_center"]])
```

Now run sparse PCA through the Rust `rupca` backend. `run_pca()` reads the
sparse PFlogPF layer plus the centering vector and writes a standard Seurat
dimensional reduction.

```{r run-pca}
lung <- run_pca(
  lung,
  assay = "RNA",
  layer = "pflogpf",
  center.key = "pflogpf_center",
  reduction.name = "scclr_pca",
  n.components = 5,
  ncv = 12
)

lung@misc$scclrR$scclr_pca$explained_variance_ratio[1:2]
```

## PCA panel

This is the R/Seurat analogue of the manuscript Figure 3 PFlogPF PCA panel.

```{r pca-plot, fig.cap = "PFlogPF PCA of Angelidis lung pseudobulk samples. Each point is one mouse-level pseudobulk sample."}
pca_df <- data.frame(
  sample = colnames(lung),
  Embeddings(lung, "scclr_pca")[, 1:2, drop = FALSE],
  age_label = lung$age_label,
  age = lung$age,
  n_cells = lung$n_cells
)
names(pca_df)[2:3] <- c("PC1", "PC2")
var <- lung@misc$scclrR$scclr_pca$explained_variance_ratio

ggplot(pca_df, aes(PC1, PC2, color = age_label, shape = age_label)) +
  geom_hline(yintercept = 0, linewidth = 0.25, color = "grey85") +
  geom_vline(xintercept = 0, linewidth = 0.25, color = "grey85") +
  geom_point(size = 3.2, stroke = 0.5) +
  geom_text_repel(
    aes(label = sample),
    size = 2.6,
    box.padding = 0.25,
    point.padding = 0.25,
    segment.color = "grey70",
    segment.linewidth = 0.2,
    min.segment.length = 0,
    max.overlaps = Inf,
    show.legend = FALSE
  ) +
  scale_color_manual(values = c(Young = "#2c7fb8", Old = "#d95f5f")) +
  scale_shape_manual(values = c(Young = 16, Old = 17)) +
  labs(
    x = sprintf("PC1 (%.1f%% variance)", 100 * var[1]),
    y = sprintf("PC2 (%.1f%% variance)", 100 * var[2]),
    color = NULL,
    shape = NULL
  ) +
  theme_classic(base_size = 11) +
  theme(legend.position = "top")
```

## edgeR old-vs-young comparison

The manuscript figure compares PC1 loadings to a supervised old-vs-young
differential expression estimate. In R, the natural replacement is `edgeR`.
For comparison, we also run the standard Seurat log-normalization workflow and
make the same PC1-loading scatterplot.

```{r seurat-default-pca}
lung_seurat <- NormalizeData(
  lung,
  assay = "RNA",
  normalization.method = "LogNormalize",
  scale.factor = 10000,
  verbose = FALSE
)
lung_seurat <- FindVariableFeatures(
  lung_seurat,
  assay = "RNA",
  nfeatures = min(2000, nrow(lung_seurat)),
  verbose = FALSE
)
lung_seurat <- ScaleData(
  lung_seurat,
  assay = "RNA",
  features = VariableFeatures(lung_seurat),
  verbose = FALSE
)
lung_seurat <- RunPCA(
  lung_seurat,
  assay = "RNA",
  features = VariableFeatures(lung_seurat),
  npcs = 5,
  reduction.name = "seurat_pca",
  verbose = FALSE
)
```

```{r edger}
group <- factor(lung$age_label, levels = c("Young", "Old"))
y <- DGEList(counts = counts, group = group)
keep <- filterByExpr(y, group = group)
y <- y[keep, , keep.lib.sizes = FALSE]
y <- calcNormFactors(y)
design <- model.matrix(~ group)
y <- estimateDisp(y, design)
fit <- glmQLFit(y, design)
qlf <- glmQLFTest(fit, coef = "groupOld")
de <- topTags(qlf, n = Inf)$table
de$gene <- rownames(de)

loadings <- data.frame(
  gene = rownames(Loadings(lung, "scclr_pca")),
  pc1_loading = Loadings(lung, "scclr_pca")[, 1]
)
merged <- merge(loadings, de, by = "gene")
if (cor(merged$pc1_loading, merged$logFC) < 0) {
  merged$pc1_loading <- -merged$pc1_loading
}
cor(merged$pc1_loading, merged$logFC)

seurat_loadings <- data.frame(
  gene = rownames(Loadings(lung_seurat, "seurat_pca")),
  pc1_loading = Loadings(lung_seurat, "seurat_pca")[, 1]
)
merged_seurat <- merge(seurat_loadings, de, by = "gene")
if (cor(merged_seurat$pc1_loading, merged_seurat$logFC) < 0) {
  merged_seurat$pc1_loading <- -merged_seurat$pc1_loading
}
cor(merged_seurat$pc1_loading, merged_seurat$logFC)
```

```{r loading-plot, fig.cap = "PFlogPF PC1 loadings compared with edgeR old-vs-young log fold changes."}
ggplot(merged, aes(pc1_loading, logFC)) +
  geom_point(aes(color = FDR < 0.05), alpha = 0.6, size = 1.2) +
  scale_color_manual(values = c(`FALSE` = "grey70", `TRUE` = "#2c7fb8"), name = "edgeR FDR < 0.05") +
  labs(x = "PFlogPF PC1 loading", y = "edgeR old-vs-young log fold change") +
  theme_classic(base_size = 11)
```

```{r seurat-loading-plot, fig.cap = "Seurat default-normalization PC1 loadings compared with edgeR old-vs-young log fold changes."}
ggplot(merged_seurat, aes(pc1_loading, logFC)) +
  geom_point(aes(color = FDR < 0.05), alpha = 0.6, size = 1.2) +
  scale_color_manual(values = c(`FALSE` = "grey70", `TRUE` = "#2c7fb8"), name = "edgeR FDR < 0.05") +
  labs(x = "Seurat default-normalization PC1 loading", y = "edgeR old-vs-young log fold change") +
  theme_classic(base_size = 11)
```

The point is not that PCA is a replacement for supervised differential
expression. Rather, PFlogPF makes the leading unsupervised geometry much less
dominated by library-depth and high-count baseline effects, so the PCA axis is
more interpretable in a sample-level RNA-seq setting.

```{r no-deps-message, eval = !has_deps, echo = FALSE}
cat(
  "This vignette needs Seurat, SeuratObject, edgeR, ggplot2, and ggrepel to execute. ",
  "The source still documents the complete workflow; pkgdown/GitHub Actions ",
  "installs those packages before rendering the site."
)
```
